Changes in Patterns of Geometric Sign Groupings as a Model of Social Dynamics Across Aurignacian Phases

### 2025-03-14 endofquarterpaper.qmd

---
title: "Changes in Groupings of Geometric Signs as a Model of Social Dynamics Across Aurignacian Periods"
author: Sophie Cooper
format: html
echo: false
execute: 
  warning: false
bibliography: references.bib
zotero: "Signbase-study"
link-citations: true
---

@@

The Upper Paleolithic period experienced a rise in more complex social organizations across Europe [@Boyd_Silk_2020] coupled with biological evolution. Another distinctive quality of the Upper Paleolithic is the large rise in artwork, both parietal and mobile. Archaeologists have frequently relied upon that artwork as a proxy for cultural difference in order to understand the shifting patterns of social organization and diversity of the Upper Paleolithic [@baker2024evidence, @Sauvet_2018, @Kuhn_Stiner_2007, @Vanhaeren_d]. Considered to be the starting point of the Upper Paleolithic, the Aurignacian technocomplex (ca. 43 - 30 k BP) is commonly linked to the dispersal of Homo sapiens across Western Eurasia, and is characterized by lithic and osseus markers [@Tejero_Grimaldi_2015; @Tartar_2012], some of which are decorated with geometric signs [@dutkiewicz2020]. In this paper, we investigate the social organization dynamics by analyzing the distribution and variation of geometric sign types found on mobile artifacts from the Upper Paleolithic in Europe across four phases of the Aurignacian period.

Social organization dynamics of the Upper Paleolithic have previously been explored using statistical analyses of artifacts to understand patterns of cultural groups. Multivariate analyses of beads from the Gravettian technocomplex suggests an east to west cline of nine cultural groups [@baker2024evidence]. Seriation analysis, PCoA, and network analysis were used to identify the groups from 134 discrete bead types recovered from both burial and occupation sites. To validate their claims of cultural groups, @baker2024evidence used a Mantel test to evaluate an isolation-by-distance hypothesis, which proposes that cultural difference can be primarily explained by geographic distance, finding that geographic distance alone did not solely account for the bead distribution. The nine geographically discrete groups encompass the regions of eastern, northwestern and central Europe, the northern and southern Iberian peninsula, southern and northern Italy, and the eastern and western Mediterranean regions.

@@

```{r}
#| label: fig-aur-table
#| fig-cap: "Table of data on Aurignacian periods [@Banks_dErrico_Zilhao_2013;  @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021; @Chu_Richter_2020; @Fletcher_Sánchez]"
library(knitr)
library(tidyverse)
library(readxl)

@@

We hypothesize that these periods, and the environmental changes across the Aurignacian, may correspond to changes in social dynamics of cultural groups, as represented by patterns in the distribution of geometric signs. We aim to address the following questions: How do grouping dynamics of geometric signs change across the four phases of the Aurignacian? How do the groupings vary in diversity, size, connectivity, and boundaries? Lastly, what do the patterns tell us about dynamics of social organization throughout the Aurignacian? We answer these questions by first identifying clusters within each time period using seriation and network analyses. We then test the strength of the clusters using modularity and perMANOVA tests. Lastly, we use Mantel tests to evaluate the null hypothesis of isolation-by-distance for each time period.

# Methods and Materials

@@

trans_unique_data <- signbase_full_clean %>% 
  filter(time_period == "transitional") %>% 
  mutate(longitude = as.character(longitude),
         latitude = as.character(latitude)) %>% 
  group_by(site_name) %>%
  summarize(across(where(is.numeric), sum)) %>% 
  left_join(lat_long_df) %>% 
  mutate(time_period = "transitional")

proto_unique_data <- signbase_full_clean %>% 
  filter(time_period == "proto_aurignacian") %>% 
  mutate(longitude = as.character(longitude),
         latitude = as.character(latitude)) %>% 
  group_by(site_name) %>%
  summarize(across(where(is.numeric), sum)) %>% 
  left_join(lat_long_df) %>% 
  mutate(time_period = "proto_aurignacian")

early_unique_data <- signbase_full_clean %>% 
  filter(time_period == "early_aurignacian") %>% 
  mutate(longitude = as.character(longitude),
         latitude = as.character(latitude)) %>% 
  group_by(site_name) %>%
  summarize(across(where(is.numeric), sum)) %>% 
  left_join(lat_long_df) %>% 
  mutate(time_period = "early_aurignacian")

evolved_unique_data <- signbase_full_clean %>% 
  filter(time_period == "evolved_aurignacian") %>% 
  mutate(longitude = as.character(longitude),
         latitude = as.character(latitude)) %>% 
  group_by(site_name) %>%
  summarize(across(where(is.numeric), sum)) %>% 
  left_join(lat_long_df) %>% 
  mutate(time_period = "evolved_aurignacian")

trans_artifact_data <- trans_unique_data %>% 
  column_to_rownames("site_name") %>% 
  dplyr::select(line:star)

proto_artifact_data <- proto_unique_data %>% 
  column_to_rownames("site_name") %>% 
  dplyr::select(line:star)

early_artifact_data <- early_unique_data %>% 
  column_to_rownames("site_name") %>% 
  dplyr::select(line:star)

evolved_artifact_data <- evolved_unique_data %>% 
  column_to_rownames("site_name") %>% 
  dplyr::select(line:star)

@@

early_unique_data <- early_unique_data %>% 
  mutate(group = case_when((site_name == "Grottes de Fonds-de-Forêt" |
                              site_name == "Riparo Bombrini"|
                              site_name == "Grotte de la Verpillière I"|
                              site_name == "Vindija Cave" |
                              site_name == "Trou al'Wesse") ~ "1",
                           (site_name == "Hohle Fels" | 
                              site_name == "Solutré" |
                              site_name == "Castanet"|
                              site_name == "Cellier" | 
                              site_name == "Blanchard") ~ "2"))

@@

@@

artifact_data <- time_unique_data_full %>% 
  column_to_rownames("id") %>% 
  select(line:star)

euc_data <- artifact_data %>% 
  mutate(sign_count = rowSums(.)) %>% 
  dplyr::select(sign_count)

euc_dist <- dist(x = euc_data, method = "euclidean")
jac_full <- vegan::vegdist(artifact_data, method = "jaccard", binary = TRUE)

euc_mantel <- vegan::mantel(euc_dist, jac_full, permutations = 1000)
p_value_abundance_correlation <- round(euc_mantel$signif, 3)
r_value_abundance_correlation <- round(euc_mantel$statistic, 3)

@@

@@

Our data was sourced from SignBase, an open-access database collection of geometric signs on mobile objects from Upper Paleolithic-era Europe. Identification and categorization of the sign types had already been performed by the creators of the SignBase database, the methods for which are described in @dutkiewicz2020, and we adhered to those categories and placements. The data set consists of `r nartifacts` artifacts found at `r nsites` sites in `r ncountries` countries, and records `r nsigntypes` geometric sign types. We excluded signs if they fell under the "other" category, due to the wide range of potential possibilities. We also excluded all data from Willendorf (only one sign present), Riparo di Fontana Nuova, Muralovka, Shanidar Cave, and Hayonim Cave (due to their extreme geographic distance from the other sites). Lastly, data from Grotte De La Princesse Pauline and Šandalja II was excluded, due to being dated much earlier than the rest of the artifacts.These exclusions resulted in `r nartifacts_clean` artifacts found at `r nsites_clean` sites in `r ncountries_clean` countries, encapsulating `r nsigntypes_clean` sign types. A map of the site distribution for each time period is shown in @fig-site-map.

```{r}
#| label: fig-site-map
#| fig-cap: "Map of sites, divided by time period"

library(rnaturalearth)
library(rnaturalearthdata)
library(ggrepel)
library(sf)
library(ggpubr)

@@

ggplot(Europe) +
  geom_sf() +
  geom_sf(data = signbase_sf) +
  coord_sf(xlim = c(-10,30), 
           ylim = c(35,53), 
           expand = FALSE) +
  geom_text_repel(data = signbase_sf,
                  aes(x = longitude ,
                      y = latitude,
                      label = site_name),
                  max.overlaps = 30,
                  size = 2) +
  theme_minimal() +
  facet_wrap(~time_period) +
  theme(axis.text = element_blank(),
        axis.title = element_blank())
```

Maximum and minimum BP dates were also provided by Signbase, and calibration was done through the rcarbon package in RStudio [@Crema_Bevan_2021]. We divided the data up into four chronological periods in accordance with the date ranges recorded in @fig-aur-table. The calibrated radiocarbon dates of the cleaned data range from approximately `r date_min_pretty` BP to `r date_max_pretty` BP, spanning an overall time range of `r date_range_pretty` years. Artifact count by time period is shown in @fig-time-count.

```{r fig.height= 5, fig.width=10}
#| label: fig-time-count
#| fig-cap: "Graph of artifact count by time period"

ggplot(signbase_full_clean) +
  aes(x = MedianBP, fill = time_period) +
  geom_bar(width = 150) +
  scale_x_reverse() +
  labs(fill = "Time period") +
  theme(legend.key.size = unit(0.5, "cm"))

@@

All our tests were run with presence/absence data. We ran a Mantel test between an Euclidean abundance distance matrix and a Jaccard distance matrix of the sign types in order to make sure that the exclusion of abundance would not significantly impact our results [@Lycett_2019], and found that it was not quite statistically significant (p-value = `r p_value_abundance_correlation`, Mantel statistic R = `r r_value_abundance_correlation`). Another Mantel test was run between a euclidean distance matrix for MedianBP and the Jaccard sign type matrix in order to determine whether there was a definitive correlation between distance in time and sign types. The Mantel test found that there was a statistically significant correlation (p-value = `r chron_p`, Mantel statistic R = `r chron_r`).

## Testing for Cultural Groupings

Seriation analysis is the arrangement of data into a linear order to reveal patterns [@Hahsler_Hornik_Buchta]. Seriation in this paper was performed using code designed to mimic the Browner-Kile seriation algorithm [@Brower_Kile_1988], in which the data is presented in a matrix diagram, and a unidimensional sequence is generated by reordering the rows and columns to group the presences along the diagonal. This is done so by calculating the mean row and column positions of presences, and arranging the rows and columns respectively. Network analysis represents the relationship between nodes [@Mills_2017]. In this network analysis, each node is a site, and each edge is a measure of sign similarity, with the direct value represented by edge weight. The results were plotted with the Fruchterman-Reingold layout [@Fruchterman_Reingold_1991], a force-directed graph layout for uniform edge lengths.

## Testing Strength of Cultural Groupings

The modularity test is community structure network detection method, and quantifies the randomness of the groupings by measuring the fraction of the edges in the network that fall within groups, minus the expected value if the edges were placed at random [@Newman_2006, @Newman_Girvan_2004]. The results are expressed in a score of 0 to 1, with 0 meaning no significant subgroups, and 1 meaning perfect groupings. The perMANOVA test quantifies the level of variation between groups versus the level of variation within groups [@Anderson_2017], with the results expressed by an R-statistic, which quantifies how much of the variation in sign distribution can be explained by group membership, and a statistical significance score (p \< 0.05), which quantifies whether the groups are significantly different.

## Testing Isolation-by-Distance Hypothesis

The isolation-by-distance framework proposes that cultural difference can be primarily explained by geographic distance. Assuming that this framework is true, we would hypothetically see patterns of cultural features of Upper Paleolithic early modern humans that are dispersed according to geography. Mantel tests are used for each time period to determine the correlation between sign distribution and geography, accomplished with use of the vegan package in RStudio [@Vegan2001]. The Mantel test is run between a geographic distance matrix and a Jaccard dissimilarity matrix. The test seeks to determine the level of correspondence between the matrices through permuational evaluation of the null distribution [@Mantel_Smouse], (here 1000 permutations) producing a Mantel R statistic and a p-value as a result. The r statistic falls between -1 and 1, with a value of -1 meaning a strong negative correlation and a value of 1 meaning a strong positive correlation, while a value of 0 means no correlation. The p-value quantifies statistical signifiance of correlation between the two matrices, with p \< 0.05 serving as the threshold for statistical significance.

@@

## Testing for Cultural Groupings

```{r}
library(ggpubr)
library(phangorn)
library(ggraph)
library(statnet)
library(ecodist)
library(ape)
library(ggrepel)
library(cowplot)
library(igraph)

@@

strength_function <- function(artifact_data, group_data){
  artifact_data <- artifact_data %>% 
    select_if(~ !is.numeric(.) || sum(.) != 0)

@@

trans_network <- function(artifact_data){
  artifact_data <- artifact_data %>% 
    select_if(~ !is.numeric(.) || sum(.) != 0)
    jac <- vegdist(artifact_data, "jaccard", binary = TRUE)
    dm <- as.matrix(jac)
    disim <- 1 - dm
    disim[disim<0.1] <- 0
    disim_net <- network(disim,
                         directed=F,
                         ignore.eval=F,
                         names.eval='weight')
    disim_net %v% 'vertex.names' <- row.names(artifact_data)
    set.seed(500)
    final_plot <- ggraph(disim_net, layout="fr") +
      geom_edge_link(aes(width = weight, 
                         alpha=weight), 
                       edge_colour = "black", 
                       show.legend=F) +
      scale_edge_width(range=c(0.5,1.5)) +
      scale_edge_colour_gradient(low = "#CCCCCC",
                                   high = "#000000") +
      geom_node_point(alpha=1, size=5) +
      geom_node_text(aes(label=row.names(artifact_data)), 
                     size=3, 
                     family= "Arial", 
                     repel=T, max.overlaps = 100) +
      theme(text= element_text(size = 15), 
            panel.background = element_rect(fill = 'white')) +
      border(color = "black", size = 0.5)
    return(final_plot)
    }
```

### Seriation

@@

For each of the time periods, seriation identifies at least two groups, one with a minimal range of signs, and one with a broader range. Diversity serves as a major dividing factor here.

### Network Analysis

@@

Network analysis shows similar groups to the ones identified in @fig-seriation, but it also shows evolving interconnectivity between groups. In the Transitional period, the network analysis shows two clearly defined groups, one consisting of Geissenklösterle, Fumane, and Pod Hradem, and the other consisting of Hohle Fels, El Castillo, and Labeko Koba, with no connections between the groups. The Proto-Aurignacian also shows two clearly defined groups, one consisting of Geissenklösterle, Spy, Vogelherd, and Grotte du Renne, and the other consisting of Mladeč, Hohlenstein-Stadel, Gatzarria, and Abri Pataud. However, it also displays some sites that are less clearly connected, with La Ferrassie being squarely in between both groups with equal connections to both, and Les Cottés lacking a connection to any of the groups. The Early Aurignacian shows more strongly interconnected groups - although the shape still suggests two main groups - with Grotte de la Verpillière I in particular having multiple strong connections to both of the groups. Lastly, the Evolved Aurignacian again shows two main groups, although group innerconnectivity has decreased, one consisting of Hohle Fels, Trou Magrite, Vogelherd, and Bockstein-Törle, and the other consisting of Gargas, La Viña, Sirgenstein Cave, Vindija Cave, and Les Rois. While interconnectivity between groups is still present, with connections between Trou Magrite, Gargas, and Vindija Cave, and Bockstein-Törle and Gargas, the strength of interconnectivity is weaker than in the Early Aurignacian.

## Groups

The results of the above tests allowed us to find two main groups of sites within each time period, as shown in @fig-group-table. The geographic distributions of the groups are shown in @fig-group-map, and the sign makeup of each group is shown in @fig-group-sign.

```{r}
#| label: fig-group-table
#| fig-cap: "Table of the sites by group"
group_table <- read_excel("data/site-groups.xlsx")

@@

## Testing Strength of Cultural Groupings

>>>>

---
title: "Changes in Patterns of Geometric Sign Groupings as a Model of Social Dynamics Across Aurignacian Phases"
author: Sophie Cooper
format: html
echo: false
execute: 
  warning: false
bibliography: references.bib
link-citations: true
---

@@

The Upper Paleolithic period experienced a rise in more complex social organizations across Europe [@Boyd_Silk_2020] coupled with biological evolution. Another distinctive quality of the Upper Paleolithic is the large rise in artwork, both parietal and mobile. Archaeologists have frequently relied upon that artwork as a proxy for cultural difference in order to understand the shifting patterns of social organization and diversity of the Upper Paleolithic [@baker2024evidence, @Sauvet_2018, @Kuhn_Stiner_2007, @Vanhaeren_d]. Considered to be the starting point of the Upper Paleolithic, the Aurignacian technocomplex (ca. 43 - 30 k BP) is commonly linked to the dispersal of Homo sapiens across Western Eurasia, and is characterized by lithic and osseus markers [@Tejero_Grimaldi_2015; @Tartar_2012], some of which are decorated with geometric signs [@dutkiewicz2020]. In this paper, we investigate the social organization dynamics by analyzing the distribution and variation of geometric sign types found on mobile objects from the Upper Paleolithic in Europe across four phases of the Aurignacian period.

Social organization dynamics of the Upper Paleolithic have previously been explored using statistical analyses of objects to understand patterns of cultural groups. Multivariate analyses of beads from the Gravettian technocomplex suggests an east to west cline of nine cultural groups [@baker2024evidence]. Seriation analysis, PCoA, and network analysis were used to identify the groups from 134 discrete bead types recovered from both burial and occupation sites. To validate their claims of cultural groups, @baker2024evidence used a Mantel test to evaluate an isolation-by-distance hypothesis, which proposes that cultural difference can be primarily explained by geographic distance, finding that geographic distance alone did not solely account for the bead distribution. The nine geographically discrete groups encompass the regions of eastern, northwestern and central Europe, the northern and southern Iberian peninsula, southern and northern Italy, and the eastern and western Mediterranean regions.

@@

```{r}
#| label: tbl-aur-table
#| tbl-cap: "Table of data on Aurignacian periods [@Banks_dErrico_Zilhao_2013;  @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021; @Chu_Richter_2020; @Fletcher_Sánchez]"
#| tbl-cap-location: bottom

library(knitr)
library(tidyverse)
library(readxl)

@@

We hypothesize that these phases, and the environmental changes across the Aurignacian, may correspond to changes in social dynamics of cultural groups, as represented by patterns in the distribution of geometric signs. We aim to address the following questions: How do grouping dynamics of geometric signs change across the four phases of the Aurignacian? How do the groupings vary in diversity, size, connectivity, and boundaries? Lastly, what do the patterns tell us about dynamics of social organization throughout the Aurignacian? We answer these questions by first identifying clusters within each phase using seriation and network analyses. We then test the strength of the clusters using modularity and perMANOVA tests. Lastly, we use Mantel tests to evaluate the null hypothesis of isolation-by-distance for each phase.

# Materials and Methods

@@

trans_df<- signbase_full_clean %>% 
  filter(time_period == "transitional") %>% 
  mutate(longitude = as.character(longitude),
         latitude = as.character(latitude))

trans_object_num <- trans_df %>% 
  group_by(site_name) %>%
  summarise(nobjects = n())

trans_unique_data <- trans_df %>% 
  group_by(site_name) %>% 
  summarize(across(where(is.numeric), sum)) %>% 
  left_join(lat_long_df) %>% 
  left_join(trans_object_num) %>% 
  mutate(time_period = "transitional")

proto_df <- signbase_full_clean %>% 
  filter(time_period == "proto_aurignacian") %>% 
  mutate(longitude = as.character(longitude),
         latitude = as.character(latitude))

proto_object_num <- proto_df %>% 
  group_by(site_name) %>%
  summarise(nobjects = n())

proto_unique_data <- proto_df %>% 
  group_by(site_name) %>%
  summarize(across(where(is.numeric), sum)) %>% 
  left_join(lat_long_df) %>% 
  left_join(proto_object_num) %>% 
  mutate(time_period = "proto_aurignacian")

early_df <- signbase_full_clean %>% 
  filter(time_period == "early_aurignacian") %>% 
  mutate(longitude = as.character(longitude),
         latitude = as.character(latitude))

early_object_df <- early_df %>% 
  group_by(site_name) %>%
  summarise(nobjects = n())

early_unique_data <- early_df %>% 
  group_by(site_name) %>%
  summarize(across(where(is.numeric), sum)) %>% 
  left_join(lat_long_df) %>% 
  left_join(early_object_df) %>% 
  mutate(time_period = "early_aurignacian")

evolved_df <- signbase_full_clean %>% 
  filter(time_period == "evolved_aurignacian") %>% 
  mutate(longitude = as.character(longitude),
         latitude = as.character(latitude))

evolved_object_df <- evolved_df %>% 
  group_by(site_name) %>%
  summarise(nobjects = n())

evolved_unique_data  <- evolved_df %>% 
  group_by(site_name) %>%
  summarize(across(where(is.numeric), sum)) %>% 
  left_join(lat_long_df) %>% 
  left_join(evolved_object_df) %>% 
  mutate(time_period = "evolved_aurignacian")

trans_artifact_data <- trans_unique_data %>% 
  column_to_rownames("site_name") %>% 
  dplyr::select(line:star) %>% 
  select_if(~ !is.numeric(.) || sum(.) != 0)

proto_artifact_data <- proto_unique_data %>% 
  column_to_rownames("site_name") %>% 
  dplyr::select(line:star) %>% 
  select_if(~ !is.numeric(.) || sum(.) != 0)

early_artifact_data <- early_unique_data %>% 
  column_to_rownames("site_name") %>% 
  dplyr::select(line:star) %>% 
  select_if(~ !is.numeric(.) || sum(.) != 0)

evolved_artifact_data <- evolved_unique_data %>% 
  column_to_rownames("site_name") %>% 
  dplyr::select(line:star) %>% 
  select_if(~ !is.numeric(.) || sum(.) != 0)

@@

early_unique_data <- early_unique_data %>% 
  mutate(group = case_when((site_name == "Grottes de Fonds-de-Forêt" |
                              site_name == "Riparo Bombrini"|
                              site_name == "Grotte de la Verpillière I"|
                              site_name == "Vindija Cave") ~ "1",
                           (site_name == "Hohle Fels" | 
                              site_name == "Solutré" |
                              site_name == "Castanet"|
                              site_name == "Cellier" | 
                              site_name == "Blanchard"|
                              site_name == "Trou al'Wesse") ~ "2"))

@@

trans_site_count <- nrow(trans_unique_data)
trans_sign_count <- ncol(trans_artifact_data)
proto_site_count <- nrow(proto_unique_data)
proto_sign_count <- ncol(proto_artifact_data)
early_site_count <- nrow(early_unique_data)
early_sign_count <- ncol(early_artifact_data)
evolved_site_count <- nrow(evolved_unique_data)
evolved_sign_count <- ncol(evolved_artifact_data)

nvogel <- nrow(signbase_full_clean %>% 
                 filter(site_name == "Vogelherd"))

artifact_count_df <- signbase_full_clean %>% 
  group_by(site_name) %>% 
  summarise(artifact_count = n())

artifact_count_median <- median(artifact_count_df$artifact_count)

@@

artifact_data <- time_unique_data_full %>% 
  column_to_rownames("id") %>% 
  dplyr::select(line:star)

object_euc_data <- time_unique_data_full%>%
  dplyr::select(id, nobjects) %>% 
  dist(method = "euclidean")

jac_full <- vegan::vegdist(artifact_data, method = "jaccard", binary = TRUE)

object_count_mantel <- vegan::mantel(object_euc_data, jac_full, permutations = 1000)
p_value_object_abundance_correlation <- round(object_count_mantel$signif, 3)
r_value_object_abundance_correlation <- round(object_count_mantel$statistic, 3)

sign_euc_data <- artifact_data %>% 
  mutate(sign_count = rowSums(.)) %>% 
  dplyr::select(sign_count) %>% 
  dist(method = "euclidean")

sign_count_mantel <- vegan::mantel(sign_euc_data, jac_full, permutations = 1000)
p_value_sign_abundance_correlation <- round(sign_count_mantel$signif, 3)
r_value_sign_abundance_correlation <- round(sign_count_mantel$statistic, 3)

@@

##chron permanova
perman_chron <- adonis2(jac_full~as.factor(time_unique_data_full$time_period), 
                    method = "jaccard",
                    sqrt.dist = TRUE)
perman_chron <- perman_chron %>% 
    data_frame() %>% 
    slice_head()

@@

Our data comes from SignBase 1.0, an open-access collection of geometric signs on mobile objects from Upper Paleolithic Europe. We used the identification and categorization of the sign types provided by SignBase's curators [@dutkiewicz2020]. The data set consists of `r nartifacts` objects found at `r nsites` sites in `r ncountries` countries, and records `r nsigntypes` geometric sign types. We excluded signs if they fell under the "other" category, due to the wide range of potential possibilities. We also excluded seven sites from the sample (only one sign present: Willendorf; extreme geographic distance from the other sites: Riparo di Fontana Nuova, Muralovka, Shanidar Cave, and Hayonim Cave; dated much later than the rest of the assemblages: Grotte De La Princesse Pauline and Šandalja II). These exclusions resulted in a sample of `r nartifacts_clean` objects found at `r nsites_clean` sites in `r ncountries_clean` countries, and `r nsigntypes_clean` sign types used in this study (@fig-site-map).

```{r fig.height= 9, fig.width=10}
#| label: fig-site-map
#| fig-cap: "Map of sites containing geometric signs on mobile objects from Upper Paleolithic Europe, sampled from the SignBase dataset, divided by Aurignacian phases"

library(rnaturalearth)
library(rnaturalearthdata)
library(ggrepel)
library(sf)
library(ggpubr)
library(cowplot)

signcount_barplot <- ggplot(signbase_full_clean) +
  aes(x = MedianBP, fill = time_period) +
  geom_bar(width = 150) +
  scale_x_reverse() +
  labs(fill = "Time period") +
  theme(legend.key.size = unit(0.5, "cm")) +
  scale_fill_discrete(guide = "none")

@@

sign_map <- ggplot(Europe) +
  geom_sf() +
  geom_sf(data = signbase_sf, aes(color = time_period)) +
  scale_colour_discrete(guide = "none")+
  coord_sf(xlim = c(-10,30), 
           ylim = c(35,53), 
           expand = FALSE) +
  geom_text_repel(data = signbase_sf,
                  aes(x = longitude ,
                      y = latitude,
                      label = site_name),
                  max.overlaps = 30,
                  size = 2, 
                  bg.color = "white",
                  bg.r = 0.25) +
  theme_minimal() +
  facet_wrap(~time_period, ) +
  theme(axis.text = element_blank(),
        axis.title = element_blank(),
        strip.text = element_text(size = 10))

plot_grid(sign_map, signcount_barplot, ncol=1, align = "v", scale = c(1, 0.8))

@@

Ages of the objects were determined by calibrating the uncalibrated ages provided by SignBase, using the rcarbon package [@Crema_Bevan_2021]. The calibrated radiocarbon dates of the sampled data range from approximately `r date_min_pretty` BP to `r date_max_pretty` BP, spanning an overall time range of `r date_range_pretty` years. We assigned each artifact to a chronological phase following the date ranges recorded in @tbl-aur-table. The Transitional phase has `r trans_site_count` sites and `r trans_sign_count` signs, the Proto-Aurignacian has `r proto_site_count` sites and `r proto_sign_count` signs, the Early Aurignacian has `r early_site_count` sites and `r early_sign_count` signs, and the Evolved Aurignacian has `r evolved_site_count` sites and `r evolved_sign_count` signs.

SignBase 1.0 is a record of objects noting only the presence or absence of each sign, without recording the quantity or sequence of individual signs on a given object. The distribution of objects across sites is highly skewed, with Vogelherd having `r nvogel`, but the median number of objects is `r artifact_count_median`. To minimize the influence of this skewed distribution, we used presence/absence data for each sign at each site, reducing the effect of variation in object count between sites. To make sure that the exclusion of abundance would not significantly impact our results, we ran a Mantel tests both between a Euclidean object abundance distance matrix and a Jaccard distance matrix of the sign types [@Lycett_2019] (p = `r p_value_object_abundance_correlation`, Mantel statistic R = `r r_value_object_abundance_correlation`), and a Eudclidean sign abundance distance matrix and a Jaccard sign type distance matrix (p = `r p_value_sign_abundance_correlation`, Mantel statistic R = `r r_value_sign_abundance_correlation`) and found that neither were statistically significant. Another Mantel test was run between a Euclidean distance matrix for dates and the Jaccard sign type matrix in order to determine whether there was a definitive correlation between distance in time and distribution of sign types, finding a statistically significant correlation (p = `r chron_p`, Mantel statistic R = `r chron_r`). This confirms that a meaningful change in sign types occurs over time.

{run permanova test for chronology}

## Cultural Group Detection Using Seriation and Network Analysis

Seriation analysis is the arrangement of data into a linear order to reveal patterns [@Hahsler_Hornik_Buchta]. We used the Browner-Kile seriation algorithm [@Brower_Kile_1988] from the seriation package [@seriation_pckge], in which a unidimensional sequence is generated by reordering the rows and columns to group the presences along the diagonal. Network analysis represents and quantifies the relationship between nodes [@Mills_2017], in this case, sites. In our network analysis, each node is a site, and each edge is a measure of sign type distribution similarity. We input the Jaccard similarity matrix for each phase, computed the network statistics, and plotted the results with the Fruchterman-Reingold layout [@Fruchterman_Reingold_1991], a force-directed graph layout for uniform edge lengths. Network analysis was performed using the vegan [@Vegan2001], statnet [@statnet_package] and igraph [@igraph_package] packages.

## Modularity and PerMANOVA for Cultural Group Strength

To evaluate the strength of the groupings found in the seriation and network analyses, we used modularity and perMANOVA tests. The modularity test is a community structure network detection method, and quantifies the randomness of the groupings by measuring the fraction of the edges in the network that fall within groups, minus the expected value if the edges were placed at random [@Newman_2006; @Newman_Girvan_2004]. The results are expressed in a score of 0 to 1, with 0 meaning no significant subgroups, and 1 meaning perfect groupings. PerMANOVA quantifies the level of variation between groups versus the level of variation within groups [@Anderson_2017]. The results are expressed by an R-statistic, which quantifies how much of the variation in sign distribution can be explained by group membership, and a p-value, which quantifies whether the groups are significantly different (we use an alpha value of 0.05 for the threshold of statistical significance). The modularity test and perMANOVA were performed using the vegan [@Vegan2001] and igraph [@igraph_package] packages.

## Effects of Geographic Distance Using Mantel Tests

The isolation-by-distance framework proposes that cultural difference can be primarily explained by geographic distance [@Lycett_2019]. If cultural difference can be explained solely by geographic distance, then we would see similarity in sign type distribution decreasing as geographic distance increases. To test this, we determined the correlation between sign distribution and geography using Mantel tests, via the vegan package [@Vegan2001]. The Mantel tests were run with a geographic distance matrix and a Jaccard similarity matrix. The test seeks to determine the level of correspondence between the matrices through permutational evaluation of the null distribution [@Mantel_Smouse], (here 1000 permutations), producing a Mantel R statistic and a p-value. The R statistic falls between -1, meaning a strong negative correlation, and 1, meaning a strong positive correlation, while a value of 0 means no correlation. The p-value quantifies statistical significance of the correlation between the two matrices, with p = 0.05 serving as the threshold for statistical significance.

@@

## Cultural Group Detection

```{r}
library(ggraph)
library(statnet)
library(ape)
library(igraph)

@@

strength_function <- function(artifact_data, group_data){

@@

trans_network <- function(artifact_data){
    jac <- vegdist(artifact_data, "jaccard", binary = TRUE)
    dm <- as.matrix(jac)
    disim <- 1 - dm
    disim[disim<0.1] <- 0
    disim_net <- network(disim,
                         directed=F,
                         ignore.eval=F,
                         names.eval='weight')
    disim_net %v% 'vertex.names' <- row.names(artifact_data)
    set.seed(500)
    final_plot <- ggraph(disim_net, layout="fr") +
      geom_edge_link(aes(width = weight, 
                         alpha=weight), 
                       edge_colour = "black", 
                       show.legend=F) +
      scale_edge_width(range=c(0.5,1.5)) +
      scale_edge_colour_gradient(low = "#CCCCCC",
                                   high = "#000000") +
      geom_node_point(alpha=1, size=5) +
      geom_node_text(aes(label=row.names(artifact_data)), 
                     size=3, 
                     family= "Arial", 
                     repel=T, max.overlaps = 100) +
      theme(text= element_text(size = 15), 
            panel.background = element_rect(fill = 'white')) +
      border(color = "black", size = 0.5)
    return(final_plot)
    }
```

@@

For each of the phases, seriation identifies at least two groups, one with a minimal range of signs, and one with a broader range. We also see that the notch and line signs are the most prevalent signs across all time periods. In the Transitional phase, we see some variety within groups, with the first consisting of Fumane, Pod Hradem, and Geissenklösterle, two of which only have the "notch" sign, and one of which has the notch and circumnotch, and the other consisting of Labeko Koba, Hohle Fels, and El Castillo, which have a broader diversity of signs. In the Proto-Aurignacian, we see an expansion in both sites and range of signs, as well as movement to new sites, with little overlap in site names. The first group consists of Abri Pataud, Gatzarria, Hohlenstein-Stadel, and Mladeč, three of which only have the notch and line signs, and one which also has a dashline sign, and the other consists of Geissenklösterle, Spy, Vogelherd, Grotte du Renne, La Ferrassie, and Les Cottés. In the second group, we see fairly consistent patterning of a diverse range of signs, although Les Cottés varies a little. In the Early Aurignacian, we again see movement to new sites with no overlap from the Proto-Aurignacian, as well as increased diversity of patterns within the groups. The first group consists of Grottes de Fonds-de-Forêt (radnotch and notch), Riparo Bombrini (notch), and Grotte de la Verpillière I (notch and line), The second consists of Solutré, Hohle Fels, Castanet, Cellier, Blanchard, and Trou al'Wesse, which contain a wide range of signs in a variety of combinations. Lastly, in the Evolved Aurignacian, we see slightly more consistent groups, again with movement to new sites with no overlap from the Early Aurignacian. The first group consists of Les Rois, Gargas, La Viña, Sirgenstein Cave, and Vindija Cave, covering the line, notch, and and obline signs. The second consists of Trou Magrite, Bockstein-Törle, Vogelherd, and Hohle Fels. Interestingly, this is the only phase to have more sites in the group with a minimal range of signs, than in the group with a more diverse range.

@@

Network analysis shows similar groups to the ones identified in @fig-seriation, but it also shows evolving inter-connectivity between groups. In the Transitional phase, the network analysis shows two clearly defined groups, one consisting of Geissenklösterle, Fumane, and Pod Hradem, and the other consisting of Hohle Fels, El Castillo, and Labeko Koba, with no connections between the groups. The Proto-Aurignacian phase also shows two clearly defined groups, one consisting of Geissenklösterle, Spy, Vogelherd, and Grotte du Renne, and the other consisting of Mladeč, Hohlenstein-Stadel, Gatzarria, and Abri Pataud. However, it also displays some sites that are less clearly connected, with La Ferrassie being squarely in between both groups with equal connections to both, and Les Cottés lacking a connection to any of the groups. The Early Aurignacian shows more strongly inter-connected groups - although the shape still suggests two main groups - with Grotte de la Verpillière I in particular having multiple strong connections to both of the groups. Lastly, the Evolved Aurignacian again shows two main groups, although group inner-connectivity has decreased, one consisting of Hohle Fels, Trou Magrite, Vogelherd, and Bockstein-Törle, and the other consisting of Gargas, La Viña, Sirgenstein Cave, Vindija Cave, and Les Rois. While inter-connectivity between groups is still present, with connections between Trou Magrite, Gargas, and Vindija Cave, and Bockstein-Törle and Gargas, the strength of inter-connectivity is weaker than in the Early Aurignacian.

In identifying the groups, we referenced both the seriation and network analyses. The results of the above tests allowed us to find two main groups of sites within each phase, as shown in @tbl-group-table. The geographic distributions of the groups are shown in @fig-group-map, and the sign makeup of each group is shown in @fig-group-sign.

```{r}
#| label: tbl-group-table
#| tbl-cap: "Table of the sites by group per phase"
#| tbl-cap-location: bottom
group_table <- read_excel("data/site-groups.xlsx")

@@

## Cultural Group Strength
